Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-05-30 jhir/latex-manuscript/manuscript.tex
update results

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states. Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%). Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%) and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%) and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} \hl{state most often transitioned to \textit{``active feedback''} (27\%) and back to} \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

In contrast, the majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%). About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}. Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements and counselor patient-centered information (18\%) distinguished \textit{``avoidant''} sequences. Both \textit{``ambivalent''} (84\%) and \textit{``avoidant''} (61\%) states most frequently transitioned to the \textit{``avoidant''} state.

>>>>

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states. \hl{We noticed that 4,724 state transitions occurred in the successful sequences including same state transitions. On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%). Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%) and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%) and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} \hl{state most often transitioned to \textit{``active feedback''} (27\%) and back to} \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

\hl{In contrast, 1,106 state transitions occurred within the unsuccessful sequences where} majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%). About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}. Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements and counselor patient-centered information (18\%) distinguished \textit{``avoidant''} sequences. Both \textit{``ambivalent''} (84\%) and \textit{``avoidant''} (61\%) states most frequently transitioned to the \textit{``avoidant''} state.
